---
title: "Mangrove Suitability Statistical Analysis"
author: "Mamed Rifqy"
date: "2026"
output: html_document
description: "Statistical analysis of mangrove rehabilitation suitability using shoreline change rate data, with spatial visualization using sf and ggplot2."
---

## Overview

This notebook performs a statistical analysis of mangrove rehabilitation suitability
across the Indragiri Hilir study area. We use shoreline change rates derived from
the GEE script to classify each polygon as Suitable, Marginal, or Not Suitable.

---

## Step 1: Load Libraries

```{r setup, message=FALSE, warning=FALSE}
library(sf)          # spatial data handling
library(dplyr)       # data wrangling
library(ggplot2)     # plotting
library(readr)       # CSV reading
library(tidyr)       # reshaping data
library(viridis)     # colour palettes

# Suppress scientific notation in output
options(scipen = 999)
```

## Step 2: Load the Data

We load the shapefile containing the rehabilitation polygons and the
CSV summary exported from the Python analysis.

```{r load-data}
# Load spatial polygons
polygons <- st_read("PRM_RIAU_2026_WGS.shp", quiet = TRUE)

# Load the suitability summary CSV
summary_df <- read_csv("mangrove_suitability_report_summary.csv")

# Preview the data
glimpse(summary_df)
```

## Step 3: Classify Suitability

Suitability is determined by the mean shoreline change rate (metres/year):
a rate below 2 m/yr is **Suitable**, between 2–10 m/yr is **Marginal**,
and above 10 m/yr is **Not Suitable**.

```{r classify}
STABLE_THRESHOLD   <- 2.0   # m/yr — Suitable threshold
MODERATE_THRESHOLD <- 10.0  # m/yr — Marginal threshold

summary_df <- summary_df %>%
  mutate(
    Suitability = case_when(
      abs(`Rate (m/yr)`) <= STABLE_THRESHOLD   ~ "SUITABLE",
      abs(`Rate (m/yr)`) <= MODERATE_THRESHOLD ~ "MARGINAL",
      TRUE                                      ~ "NOT SUITABLE"
    )
  )

# Count results
table(summary_df$Suitability)
```

## Step 4: Summary Statistics

```{r stats}
# Overall change rate statistics
summary_df %>%
  group_by(Suitability) %>%
  summarise(
    Count      = n(),
    Mean_Rate  = mean(abs(`Rate (m/yr)`), na.rm = TRUE),
    SD_Rate    = sd(abs(`Rate (m/yr)`), na.rm = TRUE),
    Min_Rate   = min(abs(`Rate (m/yr)`), na.rm = TRUE),
    Max_Rate   = max(abs(`Rate (m/yr)`), na.rm = TRUE)
  ) %>%
  arrange(Mean_Rate)
```

## Step 5: Visualise Distribution

```{r plot-distribution, fig.width=8, fig.height=5}
# Colour palette matching the suitability categories
suitability_colours <- c(
  "SUITABLE"     = "#2d5a3d",
  "MARGINAL"     = "#b45309",
  "NOT SUITABLE" = "#c0392b"
)

ggplot(summary_df, aes(x = abs(`Rate (m/yr)`), fill = Suitability)) +
  geom_histogram(bins = 20, colour = "white", linewidth = 0.3) +
  scale_fill_manual(values = suitability_colours) +
  geom_vline(xintercept = STABLE_THRESHOLD,   linetype = "dashed", colour = "#2d5a3d") +
  geom_vline(xintercept = MODERATE_THRESHOLD, linetype = "dashed", colour = "#c0392b") +
  labs(
    title    = "Distribution of Shoreline Change Rates",
    subtitle = "Dashed lines show Suitable / Marginal thresholds",
    x        = "Absolute Change Rate (m/yr)",
    y        = "Count",
    fill     = "Suitability"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    plot.title = element_text(face = "bold"),
    legend.position = "top"
  )
```

## Step 6: Spatial Join and Map

```{r spatial-map, fig.width=10, fig.height=8}
# Join suitability results back to the spatial polygons
polygons_joined <- polygons %>%
  left_join(summary_df, by = c("kelompok" = "Kelompok"))

# Plot a simple choropleth
ggplot(polygons_joined) +
  geom_sf(aes(fill = Suitability), colour = "white", linewidth = 0.3) +
  scale_fill_manual(values = suitability_colours, na.value = "grey80") +
  labs(
    title    = "Mangrove Rehabilitation Suitability Map",
    subtitle = "Indragiri Hilir, Riau — 2017–2025 Shoreline Analysis",
    fill     = "Suitability",
    caption  = "Data: Landsat 8 via Google Earth Engine"
  ) +
  theme_void(base_size = 11) +
  theme(
    plot.title    = element_text(face = "bold", size = 14),
    plot.subtitle = element_text(colour = "grey40"),
    legend.position = "bottom"
  )
```

## Step 7: Export Results

```{r export}
# Save the classified shapefile for use in GIS software
st_write(polygons_joined, "mangrove_suitability_classified.gpkg",
         driver = "GPKG", delete_dsn = TRUE, quiet = TRUE)

# Save the summary table
write_csv(summary_df, "suitability_summary_r.csv")

cat("Exported: mangrove_suitability_classified.gpkg\n")
cat("Exported: suitability_summary_r.csv\n")
```
